---
title: "Lab 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```


**Lecture**: Deep Learning (Prof. Dr. David Rügamer, Emanuel Sommer)

> Kudos to the former contributors to/creators of the lab materials Emilio Dorigatti and Tobias Weber.

Welcome to the very first lab, in which we will have fun with logistic regression.

## Exercise 1

Suppose you have five input points, $\textbf{x}_1=(0,0)^\top$, $\textbf{x}_2=(1,0)^\top$,
$\textbf{x}_3=(0,-1)^\top$, $\textbf{x}_4=(-1,0)^\top$ and $\textbf{x}_5=(0,1)^\top$, and
the corresponding classes are $y_1=y_2=y_3=0$ and $y_4=y_5=1$:

```{r eval=TRUE, echo=FALSE}
library(ggplot2)
library(latex2exp)

data = data.frame(
  x=c(0, 1, 0, -1, 0),
  y=c(0, 0, -1, 0, 1),
  t=c(r"($x_1$)", r"($x_2$)", r"($x_3$)", r"($x_4$)", r"($x_5$)"),
  class=c("0", "0", "0", "1", "1")
)

plot_data2d <- function(data, textsize = 6) {
  ggplot(data, aes(x, y)) +
    geom_point(aes(col = class)) +
    geom_text(
      aes(label = TeX(t, output = "character")), 
      nudge_x = 0.075, size = textsize, parse = TRUE
    ) + 
    labs(x = "", y = "") +
    theme_minimal()
}
plot_data2d(data)
```

Consider a logistic regression model $\hat{y}_i=\sigma\left(\alpha_0+\alpha_1x_{i1}+\alpha_2x_{i2}\right)$, with $\sigma(\cdot)$ the sigmoid function, $\sigma(x)=\left(1+e^{-x}\right)^{-1}$. What values for $\alpha_0$, $\alpha_1$ and $\alpha_2$ would result in the correct classification for this dataset? A positive label is predicted when the output of the sigmoid is larger or equal than 0.5.

***

> **Note**: do not use any formulas or automated methods to find the answer. Think for yourself. A logistic regression classifier is nothing more than a hyper-plane separating points of the two classes. If necessary, review vectors, dot-products and their geometrical interpretation in linear algebra. This applies to the following exercises, too.

***

```{r}
a0 = (
  #!hwbegin TODO fill in the value for alpha 0
  -5
  #!hwend
)

a1 = (
  #!hwbegin TODO fill in the value for alpha 1
  -10
  #!hwend
)

a2 = (
  #!hwbegin TODO fill in the value for alpha 2
  10
  #!hwend
)

# the first column is always one and is used for the "bias"
xs = matrix(c(
  1, 0, 0,
  1, 1, 0,
  1, 0, -1,
  1, -1, 0,
  1, 0, 1
), ncol = 3, byrow = T)

sigmoid = function(x) {
  #!hwbegin TODO compute and return the sigmoid transformation on x
  1 / (1 + exp(-x))
  #!hwend
}

round(
  sigmoid(xs %*% c(a0, a1, a2)),
  3
)
```

You should make sure that the last two values are close to one, and the others are close to zero.

> **Note:** There are many valid parametrization that lead to a separating hyperplane. How would you prioritize between them?

## Exercise 2

Continuing from the previous exercise, suppose now that $y_2=y_3=1$ and $y_1=y_2=y_5=0$:

```{r eval=TRUE, echo=FALSE}
data = data.frame(
  x=c(0, 1, 0, -1, 0),
  y=c(0, 0, -1, 0, 1),
  t=c(r"($x_1$)", r"($x_2$)", r"($x_3$)", r"($x_4$)", r"($x_5$)"),
  class=c("0", "1", "1", "0", "0")
)

plot_data2d(data)
```

Consider the same logistic regression model above with coefficients $\beta_0$, $\beta_1$ and $\beta_2$, how would you need to set these coefficients to correctly classify this dataset?

```{r}
b0 = (
  #!hwbegin TODO fill in the value for beta 0
  -5
  #!hwend
)

b1 = (
  #!hwbegin TODO fill in the value for beta 1
  10
  #!hwend
)

b2 = (
  #!hwbegin TODO fill in the value for beta 2
  -10
  #!hwend
)

round(sigmoid(xs %*% c(b0, b1, b2)), 3)
```

Make sure that the second and third elements are close to one, and the others close to zero.

## Exercise 3

Finally, with the same data as before, suppose that $y_1=1$ and $y_2=y_3=y_4=y_5=0$:

```{r eval=TRUE, echo=FALSE}
data = data.frame(
  x=c(0, 1, 0, -1, 0),
  y=c(0, 0, -1, 0, 1),
  t=c(r"($x_1$)", r"($x_2$)", r"($x_3$)", r"($x_4$)", r"($x_5$)"),
  class=c("1", "0", "0", "0", "0")
)

plot_data2d(data)
```

Clearly, logistic regression cannot correctly classify this dataset, since the two classes are not linearly separable (optional: prove it).

However, as we have shown in the previous exercises, it is possible to separate $x_2$ and $x_3$ from the rest, and $x_4$ and $x_5$ from the rest.

Can these two simple classifiers be composed into one that is powerful enough to separate $x_1$ from the rest?

Can we use their predictions as input for another logistic regression classifier?

Let $z_{i1}=\sigma(\alpha_0+\alpha_1x_{i1}+\alpha_2x_{i2})$ and $z_{i2}=\sigma(\beta_0+\beta_1x_{i1}+\beta_2x_{i2})$ be the output of the two logistic regression classifiers for point $i$. Then, the dataset would become:

| $i$ | $z_{i1}$ | $z_{i2}$ | $y$ |
|-----+----------+----------+-----|
| $1$ |        0 |        0 |   1 |
| $2$ |        0 |        1 |   0 |
| $3$ |        0 |        1 |   0 |
| $4$ |        1 |        0 |   0 |
| $5$ |        1 |        0 |   0 |

In graphical form:

```{r eval=TRUE, echo=FALSE}
data = data.frame(
  x=c(0, 0, 1),
  y=c(0, 1, 0),
  t=c(r"($x_1$)", r"($x_2 = x_3$)", r"($x_4 = x_5$)"),
  class=c("1", "0", "0")
)

plot_data2d(data, textsize = 4) +
  labs(x = TeX(r"($z_1$)"), y = TeX(r"($z_2$)"))
```

This sure looks linearly separable! As before, find the coefficients for a linear classifier $\hat{y}_i=\sigma\left(\gamma_0+\gamma_1z_{i1}+\gamma_2z_{i2}\right)$:

```{r}
g0 = (
  #!hwbegin TODO fill in the value for gamma 0
  5
  #!hwend
)

g1 = (
  #!hwbegin TODO fill in the value for gamma 1
  -10
  #!hwend
)

g2 = (
  #!hwbegin TODO fill in the value for gamma 2
  -10
  #!hwend
)

zs = matrix(c(
  1, 0, 0,
  1, 0, 1,
  1, 0, 1,
  1, 1, 0,
  1, 1, 0
), ncol = 3, byrow = T)

round(sigmoid(zs %*% c(g0, g1, g2)), 3)
```

Make sure that the first element is close to one, and the others close to zero.

This big classifier can be summarized as follows:

```{r}
z1 = sigmoid(xs %*% c(a0, a1, a2))
z2 = sigmoid(xs %*% c(b0, b1, b2))

yhat = sigmoid(g0 + g1 * z1 + g2 * z2)
round(yhat, 3)
```

And this is just what a neural network looks like! Each neuron is a simple linear classifier, and we just stack linear classifiers on top of linear classifiers. And we could go on and on, with many layers of linear classifiers.
